Isomorphisms as Groupoids: \\ Homotopy Type Theory with Fractional Types

### 2013-10-15 f.tex
denotations of types

Each line of the above table introduces a pair of dual
constants\footnote{where $\swapp$ and $swapt$ are self-dual.} that witness
the type isomorphism in the middle.  These are the base (non-reducible) terms
of the second, principal level of $\Pi$. Note how the above has two readings:
first as a set of typing relations for a set of constants. Second, if these
axioms are seen as universally quantified, orientable statements, they also
induce transformations of the (traditional) values. The (categorical or
homotopical) intuition here is that these axioms have computational content
because they witness isomorphisms rather than merely stating an extensional
equality.

@@

It is important to note that ``values'' and ``isomorphisms'' are completely
separate syntactic categories which do not intermix. The semantics of the
language come when these are made to interact at the ``top level'' via
\emph{application}: 
\[\begin{array}{lrcl}
\textit{top level term}, l &::=& c v
\end{array}\]

@@

The language presented above, at the type level, models a \emph{commutative
  semiring} (occasionally called a \emph{commutative rig}) where we replace
equality by isomorphism.  Semantically, $\Pi$ models a \emph{bimonoidal
  category} whose simplest example is the category of finite sets and
bijections. In that interpretation, each value type denotes a finite set of a
size calculated by viewing the types as natural numbers and each combinator
$c : b_1 \iso b_2$ denotes a bijection between the sets denoted by $b_1$ and
$b_2$. Instead of presenting such a semantics, we present a slightly more
involved semantics using discrete groupoids. In that semantics, the
denotation of each type will consist of (i) a pointed set, and (ii) a trivial
equivalence relation relating each element to itself by a \emph{path}. The
pointed sets are constructed by adjoining, to each ``regular'' set, an
element $\circ$ which topologically corresponds to the
\emph{circle}. Significantly, using the Euler characteristic as a generalized
notion of cardinality, a set containing $n$ ``regular'' elements and a circle
$\circ$ has cardinality $n$ and not $n+1$.\footnote{Generally speaking, this
  generalized cardinality based on the Euler characteristic counts the number
  of ``connected components'' of a set. Consider a set with $n$ points and
  various paths connecting them. We start by counting all the points giving
  us cardinality $n$. Then for every path, we decrease the cardinality by one
  to avoid double-counting connected elements. Then we add one again for
  every path that forms a loop because the path forming the loop did not
  decrease the number of connected components. It is easy to see that a point
  with a path that connects it to itself would have generalized cardinality
  0. Moreover the generalized cardinality of the disjoint union of two sets
  is their sum and the generalized cardinality of the product of two sets is
  their product.}

\begin{definition}[Denotation of types]
Each type $b$ is mapped to a groupoid $(\dt{b}, \circ_b, \mathit{paths}_b)$
where $\dt{b}$ is a pointed set with distinguished point $\circ_b$ and
$\mathit{paths}_b$ is a collection of paths, $v_1 \path v_2$, between elements
$v_1$ and $v_2$ of $\dt{b}$. The map is defined as follows:
\[\begin{array}{rcl@{\qquad}l}
\dt{0} &=& (\{ \circ_0 \}, \circ_0, \{ \circ_0 \path \circ_0 \}) &
          (\mbox{circle}) \\
\dt{1} &=& (\{ \bullet, \circ_1 \}, \circ_1, 
           \{ \circ_1 \path \circ_1, \bullet \path \bullet \}) &
          (\mbox{0-sphere})\\
\dt{(b_1 + b_2)} &=& \dt{b_1} \wedgesum \dt{b_2} & 
          (\mbox{wedge~sum}) \\
\dt{(b_1 * b_2)} &=& \dt{b_1} \smashproduct \dt{b_2} & 
          (\mbox{smash~product})
\end{array}\]
The empty type is mapped to the topological circle which is a point that is
connected to itself. The unit type is mapped to the topological 0-sphere
which is a pair of points with the discrete topology. The operation
$\wedgesum$ is the \emph{wedge sum} which is a regular disjoint union with a
new distinguished point $\circ_{b_1+b_2}$ and additional paths consisting of
the equivalence closure of the identifications $\circ_{b_1} \path
\circ_{b_1+b_2}$ and $\circ_{b_2} \path \circ_{b_1+b_2}$. The operation
$\smashproduct$ is the \emph{smash product} which is a regular product with a
new distinguished point $\circ_{b_1*b_2}$ and additional paths consisting of
the equivalence closure of the identifications $(x,\circ_{b_2}) \path
\circ_{b_1*b_2}$ and $(\circ_{b_1},y) \path \circ_{b_1*b_2}$ for all $x \in
\dt{b_1}$ and $y \in \dt{b_2}$.
\end{definition}

>>>>

Each line of the above table introduces a pair of dual
constants\footnote{where $\swapp$ and $\swapt$ are self-dual.} that witness
the type isomorphism in the middle.  These are the base (non-reducible) terms
of the second, principal level of $\Pi$. Note how the above has two readings:
first as a set of typing relations for a set of constants. Second, if these
axioms are seen as universally quantified, orientable statements, they also
induce transformations of the (traditional) values. The (categorical or
homotopical) intuition here is that these axioms have computational content
because they witness isomorphisms rather than merely stating an extensional
equality.

@@

It is important to note that ``values'' and ``isomorphisms'' are completely
separate syntactic categories which do not intermix. The semantics of the
language come when these are made to interact at the ``top level'' via
\emph{application}: 
\[\begin{array}{lrcl}
\textit{top level term}, l &::=& c~v
\end{array}\]

@@

The language presented above, at the type level, models a \emph{commutative
  semiring} (occasionally called a \emph{commutative rig}) where we replace
equality by isomorphism.  Semantically, $\Pi$ models a \emph{bimonoidal
  category} whose simplest example is the category of finite sets and
bijections. In that interpretation, each value type denotes a finite set of a
size calculated by viewing the types as natural numbers and each combinator
$c : b_1 \iso b_2$ denotes a bijection between the sets denoted by $b_1$ and
$b_2$. Instead of presenting such a semantics, we present a slightly more
involved semantics using discrete groupoids. In that semantics, the
denotation of each type will consist of (i) a pointed set, and (ii) a trivial
equivalence relation relating each element to itself by a \emph{path}. The
pointed sets are constructed by adjoining, to each ``regular'' set, an
element $\bot$ with a non-trivial path to itself which, topologically,
corresponds to the \emph{circle}. Significantly, using the Euler
characteristic as a generalized notion of cardinality, a set containing $n$
discrete elements (with only trivial reflexivity paths) and a circle has
cardinality $n$ and not $n+1$.\footnote{Generally speaking, this generalized
  cardinality based on the Euler characteristic counts the number of
  ``connected components'' of a set. Consider a set with $n$ points and
  various paths connecting them. We start by counting all the points giving
  us cardinality $n$. Then for every path, we decrease the cardinality by one
  to avoid double-counting connected elements. Then we add one again for
  every path that forms a loop because the path forming the loop did not
  decrease the number of connected components. It is easy to see that a point
  with a path that connects it to itself would have generalized cardinality
  0. Moreover the generalized cardinality of the disjoint union of two sets
  is their sum and the generalized cardinality of the product of two sets is
  their product.}

\begin{definition}[Denotation of types]
Each type $b$ is mapped to a groupoid $(\dt{b}, \bot_b, \mathit{paths}_b)$
where~$\dt{b}$ is a pointed set with distinguished point $\bot_b$ and
$\mathit{paths}_b$ is a collection of paths, $v_1 \path v_2$,
between elements $v_1$ and $v_2$ of $\dt{b}$. (We omit the trivial paths for
reflexivity and only include the non-trivial ``generating'' paths as
suggested by the notation of \emph{higher-order inductive types}.)
\[\begin{array}{rcl@{\qquad}l}
\dt{0} &=& (\{ \bot_0 \}, \bot_0, \{ \bot_0 \path \bot_0 \}) &
           (\mbox{circle}) \\
\dt{1} &=& (\{ \bullet, \bot_1 \}, \bot_1, \{ \bot_1 \path \bot_1 \}) &\\
\dt{(b_1 + b_2)} &=& \dt{b_1} \wedgesum \dt{b_2} & 
           (\mbox{wedge~sum}) \\
\dt{(b_1 * b_2)} &=& \dt{b_1} \smashproduct \dt{b_2} & 
           (\mbox{smash~product})
\end{array}\]
The empty type is mapped to the topological circle which is a point that is
connected to itself. The unit type is mapped to a pointed singleton with a
distinguished element connected to itself with a non-trivial loop. The
operation $\wedgesum$ is the \emph{wedge sum} which corresponds to the
disjoint union with an additional path $\leftv{\bot_{b_1}} \path
\rightv{\bot_{b_2}}$ identifying the two distinguished elements of the
component spaces. The new cluster is the distinguished element of the
composite space. The operation $\smashproduct$ is the \emph{smash product}
which is the regular set product with additional paths connecting
$(x,\bot_{b_2})$ and $(\bot_{b_1},y)$ for all $x \in \dt{b_1}$ and $y \in
\dt{b_2}$. The new cluster of identified distinguished elements is the new
distinguished element of the composite space.
\end{definition}
